---
title: "Student performance analysis"
author: "Fares Mohamed"
date: "2024-10-11"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## The Analysis process of The Student performance data.

### First we call the libraries needed for the Analysis.
```{r}
library(ggplot2)
library(tidyverse)
library(tidyr)
library(dplyr)
library(scales)
```

### Then we call the Dataset that we  will preform the analysis on.
```{r}
df <- read.csv("C:/Users/farem/Desktop/Group project/cleaned_StudentPerformanceFactors.csv")
```
### then we View our dataset
```{r}
view(df)
```
## Now we can start the visualising our data for insights.
### the points we are going to visulize here are: 

* School type with education level by gender
* Cumulative Exam Scores by hours studied 
* Hours studied by exam scores over time
* Attendance by exam scores
* Parental and environmental influence
* Family income vs exam scores 
* Exam Scores by Gender 

### 1. School type with education level by gender
```{r}
ggplot(df, aes(x = Parental_Education_Level, y = Hours_Studied, fill = Gender)) +
  geom_bar(stat = "identity", position = "dodge") +
  facet_wrap(~ School_Type) +
  labs(title = "Distribution of School Type by Education Level and Gender",
       x = "Parental Education Level",
       y = "Hours_Studied",
       fill = "Gender") +
  theme_minimal()
```

### 2. Cumulative Exam Scores by hours studied
```{r}
ggplot(df, aes(x = Hours_Studied, y = Exam_Score, fill = Gender, color = Gender)) +
  stat_summary(fun = mean, geom = "area", alpha = 0.3, position = "identity") +
  stat_summary(fun = mean, geom = "line", linewidth = 1.5) +
  stat_summary(fun = mean, geom = "point", size = 2) +
  labs(title = "Average Exam Scores by Hours Studied by Gender", 
       x = "Hours Studied", 
       y = "Average Exam Scores") +
  scale_fill_manual(values = c("Female" = "blue", "Male" = "orange")) +
  scale_color_manual(values = c("Female" = "blue", "Male" = "orange")) +
  theme_minimal() +
  theme(legend.title = element_blank())
```

### 3. Average Exam Scores by Hours Studied Over Time

```{r}
ggplot(df, aes(x = Hours_Studied, y = Exam_Score)) +
  stat_summary(fun = mean, geom = "line", color = "blue") +
  labs(title = "Average Exam Scores by Hours Studied Over Time", x = "Hours Studied", y = "Average Exam Scores") +
  theme_minimal()
```

### 4. Exam Scores by Attendance

```{r}
ggplot(df, aes(x = Attendance, y = Exam_Score)) +
  stat_summary(fun = mean, geom = "line", color = "blue") +
  labs(title = "Exam Scores by Attendance", x = "Attendance", y = "Exam Scores") +
  theme_minimal()
```

### 5. Parental and environmental influence

```{r}
plot_data <- df %>%
  group_by(Parental_Involvement) %>%
  summarize(
    High_Motivation = sum(Exam_Score[Motivation_Level == "High"]),
    Medium_Motivation = sum(Exam_Score[Motivation_Level == "Medium"]),
    Low_Motivation = sum(Exam_Score[Motivation_Level == "Low"])
  ) %>%
  tidyr::gather(key = "Motivation", value = "Exam_Score", -Parental_Involvement)

# parental and environmental influence

ggplot(plot_data, aes(x = Parental_Involvement, y = Exam_Score, fill = Motivation)) +
  geom_bar(stat = "identity") +
  labs(title = "Parental and Environmental Influence", 
       subtitle = "Parental Involvement", 
       x = "Parental Involvement", 
       y = "Exam Score") +
  theme_minimal()
```

### 6. Family income vs exam scores

```{r}
pie_data <- df %>%
  group_by(Family_Income, Gender) %>%
  summarise(Count = n(), .groups = 'drop') %>%
  mutate(Percentage = Count / sum(Count) * 100)

#create pie chart
ggplot(pie_data, aes(x = "", y = Percentage, fill = interaction(Family_Income, Gender))) +
  geom_bar(stat = "identity", width = 1, color = "black") +
  coord_polar(theta = "y") +
  labs(title = "Distribution of Exam Scores by Family Income and Gender", fill = "Income and Gender") +
  theme_minimal() +
  theme(axis.text.x = element_blank(), 
        axis.title.x = element_blank(),
        axis.title.y = element_blank()) +
  geom_text(aes(label = paste0(round(Percentage, 1), "%")), 
            position = position_stack(vjust = 0.5), 
            color = "black") 
```

### 7. Exam Scores by Gender

```{r}
total_scores <- df %>%
  group_by(Gender) %>%
  summarise(Total_Score = sum(Exam_Score, na.rm = TRUE), .groups = 'drop') %>%
  mutate(Percentage = Total_Score / sum(Total_Score) * 100)

# Create the pie chart
ggplot(total_scores, aes(x = "", y = Percentage, fill = Gender)) +
  geom_bar(stat = "identity", width = 1, color = "white") +
  coord_polar(theta = "y") +
  labs(title = "Exam Scores by Gender", fill = "Gender") +
  theme_minimal() +
  theme(axis.text.x = element_blank(), 
        axis.title.x = element_blank(),
        axis.title.y = element_blank()) +
  geom_text(aes(label = paste0(round(Percentage, 1), "%")), 
            position = position_stack(vjust = 0.5), 
            color = "white")
```
